---
title: 'Module 1: Working with spatial data'
author: "Brenna Kelly"
date: "2024-07-16"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

Make sure you have the following packages installed:  

```{r message=FALSE, warning=FALSE}

# install.packages(c("tmap", "ggpubr", "ggplot2", 
#                    "ggspatial", "prettymapr",
#                    "tidyr", "stringr", "dplyr"))

library(sf)
library(tmap)
library(dplyr)

```

  
## Spatial data are just `simple features`

The `sf` (simple features) package provides a standardized way to encode spatial vector data. With this resource, we can:  
- read and write spatial data  
- perform geometric operations  
- transform and project our spatial data.  

Let's get read to read in our first set of spatial data:  

```{r}

list.files("data/guerry")

```

`Guerry` contains socioeconomic data for 1830 France, collected by Andre-Michel Guerry as "Morality Statistics". There are four types of files in the `guerry` folder, in addition to the documentation:  
- `dbf`: stores the feature attributes in a tabular format  
- `prj`: contains the projection and coordinate reference system  
- `shp`: contains the geometry  
- `shx`: indexes the geometry  

*We read in the data using `st_read` with the file path for the `shp` file. Do this for the Guerry data, and assign it to an object named `guerry`. Check its class and the first few rows of data.  *

```{r}



```

We see in the dataframe (and can confirm visually) that these are polygons.

```{r echo = FALSE}

tm_shape(guerry) +
  tm_polygons()

```

Geometry can be as simple as a point, a line, a polygon, a multipolygon, or a collection of these things.

```{r, echo = FALSE, warning = FALSE}

# one point
guerry_center <- st_centroid(guerry[1, ])
# multiple points
guerry_centers <- st_centroid(guerry)

# one line
guerry_centers <- st_centroid(guerry)
guerry_line <- guerry_centers[1:2, ] |>
  summarise(do_union = TRUE) |>
  st_cast("MULTILINESTRING")
# multiple lines
guerry_lines <- guerry_centers[1:6, ] |>
  summarise(do_union = TRUE) |>
  st_cast("MULTILINESTRING")

# one polygon
# st_geometry(guerry[1, ])
# multiple polygons
seine <- guerry |>
  filter(Dprtmnt %in% c("Seine", "Seine-et-Marne",
                        "Seine-et-Oise"))

# one multipolygon
finistere <- guerry |>
  filter(Dprtmnt %in% c("Finistere"))
# multiple multipolygon
bretagne <- guerry |>
  filter(Dprtmnt %in% c("Cotes-du-Nord", "Finistere",
                        "Morbihan", "Ille-et-Vilaine"))

```

```{r echo = FALSE}

par(mfrow = c(2, 4))
plot(st_geometry(guerry_center))
plot(st_geometry(guerry_line))
plot(st_geometry(guerry[1, ]))
plot(st_geometry(finistere))

plot(st_geometry(guerry_centers))
plot(st_geometry(guerry_lines))
plot(st_geometry(seine))
plot(st_geometry(bretagne))

```

## Coordinate Reference Systems

Importantly, every simple feature needs a coordinate reference system (CRS), because while we use two-dimensional coordinates to describe where things are located, Earth is not flat (surprise!). The CRS defines how the Earth is "flattened" for mapping. We could spend several lectures discussing coordinate reference systems, but the key takeaways for our purposes are:  
1. CRS contains the rules for projection and the datum being used (origin point on Earth's surface, unit of coordinates, and orientation).  
2. Difference reference systems exist, and they are designed to optimize different features (shape, distance, area). Your CRS choice should match your analysis needs.  
3. When using multiple datasets, your CRS needs to match.  

- Use `st_crs()` to check the coordinate reference system of the `guerry` data.  

```{r}


```

The `guerry` dataset uses a projected coordinate system, which is optimized for a particular location to minimize visual distortion. Compare against the unprojected WGS84, NAD83 (projected for North America), and the Mercator-projected WGS84. We use `st_transform(, crs = )` to reproject.

```{r}

ntf <- tm_shape(guerry) +
  tm_polygons() +
  tm_layout(title = "Nouvelle Triangulation \nFrancaise (Paris)", title.size = 1)

guerry_wgs <- st_transform(guerry, crs = 4326)
wgs <- tm_shape(guerry_wgs,) +
  tm_polygons() +
  tm_layout(title = "World Geodetic \nSystem 1984, \nUnprojected", title.size = 1)

guerry_aea <- st_transform(guerry, crs = st_crs("+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=37.5 +lon_0=-96 +ellps=GRS80 +datum=NAD83"))
aea <- tm_shape(guerry_aea) +
  tm_polygons() +
  tm_layout(title = "North American \nDatum 1983", title.size = 1)

guerry_merc <- st_transform(guerry, crs = 3395)
merc <- tm_shape(guerry_merc) +
  tm_polygons() +
  tm_layout(title = "WGS84 with \nMercator Projection", title.size = 1)

tmap_arrange(ntf, wgs, aea, merc, ncol = 2, nrow = 2)
dev.off()

```


## Creating simple features

Often you'll find coordinates in a data set, but it hasn't been converted to a simple feature.

- Read in the filed located at `data/spirals/spirals.csv` and call it `spiral`.  

```{r}


```

We could simply plot the x- and y-coordinates, but without a CRS, we cannot combine this with any other information, and we won't get much further use out of the data.

- Plot the x- and y-coordinates with the `plot` function.  

```{r}


```

To create a simple feature, we use the `st_as_sf` function. This requires us to specify the column names for the coordinates (e.g., x and y, longitude and latitude) and the EPSG code for the coordinate reference system (if you don't know what it should be, it's easily google-able).

- Using `st_as_sf`, transform the spiral object using `crs = 4326`. Call this new object spiral_sf  

```{r}


```

Note that if we did not assign a `crs` to the object, the code would still run. However, if we decide to project the data later on, we will get an error saying we cannot transform an object with a missing CRS We would need to assign a CRS to the object with `st_crs(spiral_sf) <- 4326`. Best practice is to always have a CRS for your sf objects.  

Now that we have our `spiral_sf`, we can plot this feature:

```{r}

tm_shape(spiral_sf) +
  tm_dots(col = "skyblue", size = 0.5, alpha = 0.75)

```

To save an `sf` object, we use the `st_write` function. Note that long column names may be abbreviated, so it may look different when you read it back it. Unlike `write.csv`, `st_write` will give you an error if you try to overwrite an existing file (use `append = FALSE` to override this).

- Create a new column in the spiral_sf object called `point_number`. This is a long name and may be abbreviated when we use `st_write()` to save the sf object.  

```{r, warning=FALSE}


```

We can also create simple features based on existing `sf` objects. We'll demonstrate using a counties in Utah and points contained within counties.

```{r}

utah <- st_read("data/utah/utah.shp")
utah_points <- st_read("data/utah/utah_points.shp")

```

```{r, echo = FALSE}

tm_shape(utah) +
  tm_polygons(col = "white") +
  tm_shape(utah_points) +
  tm_dots(col = "red", size = 0.25)

```

- Find the geometric centroid of Salt Lake City using `st_centroid`. Then using `st_buffer`, create a buffer using around the centroid (the units of the distance are determined by the CRS, which you can check using `st_crs(slc_center)$units_gdal`). The buffer is of the type `polygon`.  

```{r}

slc <- subset(utah, NAME == "Salt Lake County, Utah")

```

```{r echo = FALSE}

tm_shape(slc) +
  tm_polygons() +
  tm_shape(slc_center_buff) +
  tm_polygons(col = "seagreen") +
  tm_shape(slc_center) +
  tm_dots(shape = 8, size = 2)

```

- Use the same functions on all counties in Utah. Using `st_filter`, keep only the `utah_points` which fall inside these buffers.  

```{r}


```

To examine the results, we'll make a plot. The simplest way to plot a simple feature is using `plot()` and `st_geometry`. If you use `plot(sf_data)`, you will get a map for every column in the data. `st_geometry` allows you to just access the geometry.

- Plot the Utah counties, their centroids, and the buffers around the centroids. Try to add the points which only lie within the buffers.  

```{r}


```


## Exercise

Let's practice working with spatial data using the UK pubs dataset.  
- Read in the `uk.shp` shapefile and `pubs.csv` file.  

```{r}

uk <- st_read("data/uk_pubs/uk.shp")
pubs <- read.csv("data/uk_pubs/pubs.csv")

```

- Convert the `pubs` dataframe to an sf object, with `crs = 4277`, the geographic coordinates for the UK. 

```{r}

```

- See if there are differences in the CRS for `uk` and `pubs` (you can check manually and/or visually with a plot). If needed, transform the `pubs` to the `uk` CRS.  

```{r}

```

- Choose one district to `st_subset` (e.g., LAD24NM == Lancaster, Camden, Cardiff, Sheffield).  

```{r}

```

- Create a 5-km buffer around the area (the units are meters, so use `dist = 5000`).  

```{r}

```

- Using the `st_filter()` function, can you determine whether there are more pubs inside the area or within 5 km of the area?  

```{r}

```







## Exercise Solution

```{r}

uk <- st_read("data/uk_pubs/uk.shp")
pubs <- read.csv("data/uk_pubs/pubs.csv")

# make spatial
pubs <- st_as_sf(pubs, coords = c("lon", "lat"),
                     crs = 4277)

```

```{r}

# check for CRS mismatch
plot(st_geometry(uk))
plot(st_geometry(pubs), col = "red")

plot(st_geometry(uk))
plot(st_geometry(pubs), add = TRUE, col = "red")

st_crs(pubs) == st_crs(uk) # FALSE

```

```{r}

# transform
pubs <- st_transform(pubs, crs = st_crs(uk))

# plot with base r
plot(st_geometry(uk))
plot(st_geometry(pubs), add = TRUE, col = "red")

# plot with tmap
tm_shape(uk) +
  tm_polygons(col = "blue") +
  tm_shape(pubs) +
  tm_dots(col = "red")

```

```{r}

# are there more pubs within Maidstone or within 5 km of Maidstone?
maidstone <- subset(uk, LAD24NM == "Maidstone")

# buffer around polygon
maidstone_buffer <- st_buffer(maidstone, dist = 5000)
# how many pubs are *near* the area
near_maidstone_pubs <- st_filter(pubs, maidstone_buffer)

# how many pubs are in the area
maidstone_pubs <- st_filter(pubs, maidstone)

pubs_near_maidstone = (nrow(near_maidstone_pubs) - nrow(maidstone_pubs)) # we want to exclude pubs inside the area
pubs_in_maidstone = nrow(near_maidstone_pubs)

# are there more pubs within
pubs_in_maidstone / pubs_near_maidstone

# let's visually inspect
plot(st_geometry(maidstone_buffer), col = "olivedrab3")
plot(st_geometry(maidstone), col = "olivedrab4", add = TRUE)
plot(st_geometry(near_maidstone_pubs), col = "darkgoldenrod3", pch = 17, add = TRUE)
plot(st_geometry(maidstone_pubs), col = "darkgoldenrod1", pch = 17, add = TRUE)

# with tmaps
tm_shape(maidstone_buffer) +
  tm_polygons(col = "olivedrab3", lwd = 0) +
  tm_shape(maidstone) +
  tm_polygons(col = "olivedrab4", lwd = 0, alpha = 0.5) +
  tm_shape(near_maidstone_pubs) +
  tm_dots(col = "darkgoldenrod3", shape = 17, size = 0.5, alpha = 0.5) +
  tm_shape(maidstone_pubs) +
  tm_dots(col = "darkgoldenrod1", shape = 17, size = 0.5, alpha = 0.5)


```

There are twice as many pubs (2.03-times) in Maidstone as there are within 5 km of Maidstone.

